Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model

### 2012-01-19 Multiresolution_framework/ltx/manuscript.tex
aded bullet points for the EM algorithm

By assuming a linear activation function of the form
\begin{equation}
	f(v_t(\mathbf{r})) = \varsigma v_t(\mathbf{r})
\end{equation}
the state transition equation~\eqref{eq:ApproxDiscreteTimeModel4} simplifies to
\begin{equation}\label{eq:ApproxDiscreteTimeModel_Linear}
	\mathbf{x}_{t+1} = 
	\xi \mathbf{x}_t + 
	\varsigma T_s \mathbf{\Lambda}_{x}^{-1} \int_{\Omega}\boldsymbol\mu\left(\mathbf{r}\right)\int_\Omega { 
	    \boldsymbol\theta^\top\boldsymbol\lambda\left(\mathbf{r}-\mathbf{r}'\right)
	    \boldsymbol\mu^\top\left(\mathbf{r}'\right) 
	\, \mathrm{d}\mathbf{r}'\mathrm{d}\mathbf{r}} \mathbf{x}_t
	+ \mathbf{w}_t.
\end{equation}
Now we can further simplify the system by defining
\begin{align}
	\label{eq:Lambdatheta}
	 \mathbf{\Lambda}_{\theta} &\triangleq \int_{\Omega}\boldsymbol\mu\left(r\right) \int_\Omega { 
		   \boldsymbol\theta^\top\boldsymbol\lambda\left(r-r'\right)
		    \boldsymbol\mu^\top\left(r\right)\ \mathrm{d}r'\mathrm{d}r}
\end{align}
to give the state transition equation
\begin{align}\label{eq:StateEquation}
 \mathbf x_{t+1} &=\mathbf A(\boldsymbol \theta) \mathbf x_t+ \mathbf w_t\\
\label{eq:A_theta}
 \mathbf A(\boldsymbol \theta) &= T_s\varsigma\mathbf{\Lambda}_{x}^{-1}\mathbf{\Lambda}_{\theta}+\xi\mathbf I.
\end{align}

@@

A derivation of the lower bound that is based on \cite{Bishop2006} is provided in this Appendix. Let the entire sequence of observations is denoted by $\mathbf Y$, and the sequence of hidden variables by $\mathbf X$ and the parameter set by $\boldsymbol\Theta$. We want to maximise the likelihood function which is difficult to maximise directly

@@

The EM algorithm alternates between maximising the lower bound with respect to the distribution $q(\mathbf X)$ and the parameter set $\boldsymbol\Theta$, respectively, while holding the other fixed. Suppose that the current estimate of the parameter set is $\boldsymbol\Theta'$, in the E-step the lower bound is maximised with respect to the distribution $q(\mathbf X)$ while $\boldsymbol\Theta'$ is fixed. Since the value of $p(\mathbf Y;\boldsymbol\Theta')$ is independent of $q(\mathbf X)$ the maximum of the lower bound occurs when the Kullback-Liebler distance disappears or in other words when $q(\mathbf X)$ is exactly the conditional distribution $p(\mathbf X \mid \mathbf Y;\boldsymbol\Theta)$. At this point, the lower bound becomes equal to the log-likelihood function. The maximum in the M-step is obtained by maximising the lower bound with respect to the parameter set while the distribution $q(\mathbf X)$ is held fixed. Since the distribution $q(\mathbf X)$ is computed using the old parameter estimates and kept fixed during the M-step, it is not equal to $p(\mathbf X \mid \mathbf Y;\boldsymbol\Theta)$, resulting in a non-zero KL distance. Therefore, the growth of the log-likelihood function is greater than the growth of the lower bound. Substituting $p(\mathbf X \mid \mathbf Y;\boldsymbol\Theta')$ for $q(\mathbf X)$ in \eqref{eq:EM-decomposition1} gives
\begin{align}
 \mathcal{L}(q,\boldsymbol\Theta)&=\int_{\mathbf X} p(\mathbf X\mid\mathbf Y;\boldsymbol\Theta')\ln p(\mathbf X,\mathbf Y;\boldsymbol\Theta)~d\mathbf X-\int_{\mathbf X} p(\mathbf X\mid\mathbf Y;\boldsymbol\Theta')\ln p(\mathbf X\mid\mathbf Y;\boldsymbol\Theta')~d\mathbf X \nonumber \\
&= \mathbf E_{\boldsymbol\Theta'}\left[ \ln p(\mathbf X,\mathbf Y;\boldsymbol\Theta)\right] +\mathrm{const.}\nonumber\\
&=\mathcal{Q}(\boldsymbol\Theta,\boldsymbol\Theta')+\mathrm{const.}~,\label{eq:EM-ComDaLogLike}
\end{align}
where $ \mathbf E_{\boldsymbol \Theta'}\left[ .\right] $ denotes expectation taken with respect to the marginal distribution of the states conditioned on the observed field and the current estimate of the parameter set, i.e.  $p(\mathbf X\mid\mathbf Y;\boldsymbol \Theta'$). The constant term is the negative entropy of $q(\mathbf X)$ and is independent of $\boldsymbol\Theta$.   It becomes clear from \eqref{eq:EM-ComDaLogLike} that, the M-step essentially maximises the expectation of the complete data log-likelihood.

>>>>

By assuming a linear activation function of the form
	f(v_t(\mathbf{r})) = \varsigma v_t(\mathbf{r})
the state transition equation~\eqref{eq:ApproxDiscreteTimeModel4} simplifies to
\begin{equation}\label{eq:ApproxDiscreteTimeModel_Linear}
	\varsigma T_s \mathbf{\Lambda}_{x}^{-1} \int_{\Omega}\boldsymbol\mu\left(\mathbf{r}\right)\int_\Omega { 
	    \boldsymbol\theta^\top\boldsymbol\lambda\left(\mathbf{r}-\mathbf{r}'\right)
	    \boldsymbol\mu^\top\left(\mathbf{r}'\right) 
	\, \mathrm{d}\mathbf{r}'\mathrm{d}\mathbf{r}} \mathbf{x}_t
	+ \mathbf{w}_t.
Now we can further simplify the system by defining
	\label{eq:Lambdatheta}
	 \mathbf{\Lambda}_{\theta} &\triangleq \int_{\Omega}\boldsymbol\mu\left(r\right) \int_\Omega { 
		    \boldsymbol\mu^\top\left(r\right)\ \mathrm{d}r'\mathrm{d}r}
to give the state transition equation
\begin{align}\label{eq:StateEquation}
 \mathbf x_{t+1} &=\mathbf A(\boldsymbol \theta) \mathbf x_t+ \mathbf w_t\\
\label{eq:A_theta}
 \mathbf A(\boldsymbol \theta) &= T_s\varsigma\mathbf{\Lambda}_{x}^{-1}\mathbf{\Lambda}_{\theta}+\xi\mathbf I.
\parham{
 \begin{itemize}
	\item when likelihood is difficult to maximise directly (when we have hidden variable) so we maximise the joint likelihood function 
	\item EM maximises the lower bound and show what exactly this lower bound is (shown in Appendix).
	\item E-step for state-space model is the smoothing problem
	\item so when we have joint state parameter estimation for stae-space models  we only need to do two things: 1- use smoother for the E-step 2- construct the lower bound which might not be straightforward even in linear case
\end{itemize}}

@@

A derivation of the lower bound on the likelihood function that is based on \cite{Bishop2006} is provided in this Appendix. Let the entire sequence of observations is denoted by $\mathbf Y$, and the sequence of hidden variables by $\mathbf X$ and the parameter set by $\boldsymbol\Theta$. We want to maximise the likelihood function which is difficult to maximise directly

@@

The EM algorithm alternates between maximising the lower bound with respect to the distribution $q(\mathbf X)$ and the parameter set $\boldsymbol\Theta$, respectively, while holding the other fixed. Suppose that the current estimate of the parameter set is $\boldsymbol\Theta'$, in the E-step the lower bound is maximised with respect to the distribution $q(\mathbf X)$ while $\boldsymbol\Theta'$ is fixed. Since the value of $p(\mathbf Y;\boldsymbol\Theta')$ is independent of $q(\mathbf X)$ the maximum of the lower bound occurs when the KL distance disappears or in other words when $q(\mathbf X)$ is exactly the conditional distribution $p(\mathbf X \mid \mathbf Y;\boldsymbol\Theta)$. At this point, the lower bound becomes equal to the log-likelihood function. Note in statespace model the E-step is equiavalent to the smoothing problem. The maximum in the M-step is obtained by maximising the lower bound with respect to the parameter set while the distribution $q(\mathbf X)$ is held fixed. Since the distribution $q(\mathbf X)$ is computed using the old parameter estimates and kept fixed during the M-step, it is not equal to $p(\mathbf X \mid \mathbf Y;\boldsymbol\Theta)$, resulting in a non-zero KL distance. Therefore, the growth of the log-likelihood function is greater than the growth of the lower bound. Substituting $p(\mathbf X \mid \mathbf Y;\boldsymbol\Theta')$ for $q(\mathbf X)$ in \eqref{eq:EM-decomposition1} gives
\begin{align}
 \mathcal{L}(q,\boldsymbol\Theta)&=\int_{\mathbf X} p(\mathbf X\mid\mathbf Y;\boldsymbol\Theta')\ln p(\mathbf X,\mathbf Y;\boldsymbol\Theta)~d\mathbf X-\int_{\mathbf X} p(\mathbf X\mid\mathbf Y;\boldsymbol\Theta')\ln p(\mathbf X\mid\mathbf Y;\boldsymbol\Theta')~d\mathbf X \nonumber \\
&= \mathbf E_{\boldsymbol\Theta'}\left[ \ln p(\mathbf X,\mathbf Y;\boldsymbol\Theta)\right] +\mathrm{const.}\nonumber\\
&=\mathcal{Q}(\boldsymbol\Theta,\boldsymbol\Theta')+\mathrm{const.}~,\label{eq:EM-ComDaLogLike}
\end{align}
where $ \mathbf E_{\boldsymbol \Theta'}\left[ .\right] $ denotes expectation taken with respect to the marginal distribution of the states conditioned on the observed field and the current estimate of the parameter set, i.e.  $p(\mathbf X\mid\mathbf Y;\boldsymbol \Theta'$). The constant term is the negative entropy of $q(\mathbf X)$ and is independent of $\boldsymbol\Theta$.   It becomes clear from \eqref{eq:EM-ComDaLogLike} that, the M-step essentially maximises the expectation of the complete data log-likelihood.
